The illusory promise of the Aligned Rank Transform

### 2025-07-08 text/30-assumptions.qmd
small corrections

While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Naturally, as the shapes become more similar, the problem diminishes. In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.

>>>>

While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html#log-normal-distributions-with-equal-variances) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Naturally, as the shapes become more similar, the problem diminishes. In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.

### 2025-07-08 text/62-Results-AcrossExpDesigns.qmd
small corrections

### Type I errors across experimental designs {#extra}
We investigates the generalizability of the aforementioned results to other experimental designs involving two or three factors. We assess five out of the six ratio scales examined earlier. We exclude the Cauchy distribution and replace it by a 5-level ordinal scale with flexible thresholds.

>>>>

### Type I errors across experimental designs {#extra}
We investigate the generalizability of the aforementioned results to other experimental designs involving two or three factors. We assess five out of the six ratio scales examined earlier. We exclude the Cauchy distribution and replace it by a 5-level ordinal scale with flexible thresholds.

### 2025-07-08 text/63-Results-UnequalVariance.qmd
small corrections

**Main effects**. We begin by investigating how the four methods detect main effects on $X_1$. The interpretation of results in this scenario is challenging due to differences in the original populations -- although they share the same means (and medians), their variances differ. Depending on the null hypothesis of interest, conclusions may vary. Complicating matters further, non-linear transformations can result in distributions with differing means and medians. Consequently, the choice of statistical method may lead to different outcomes; one method may be sensitive to mean differences while others to median differences.

>>>>

**Main effects**. We begin by investigating how the four methods detect main effects on $X_1$. The interpretation of results in this scenario is challenging due to differences in the original populations --- although they share the same means (and medians), their variances differ. Depending on the null hypothesis of interest, conclusions may vary. Complicating matters further, non-linear transformations can result in distributions with differing means and medians. Consequently, the choice of statistical method may lead to different outcomes; one method may be sensitive to mean differences while others to median differences.

### 2025-07-08 text/66-Results-Interactions.qmd
small corrections

- *RNK.* Error rates explode when both main effects exceed a certain level (e.g., when $a_1, a_2 \ge 4$ and $n = 20$). This problem arises from the way the rank transformation distorts interaction effects (see @fig-interactions-rank). For all continuous distributions, we observe that error rates exhibit the exact same patterns, since ranks are scale-invariant.

@@

- *ART.* It maintains correct error rates as long as population distributions are normal. For all other distributions, error rates increase rapidly as effect sizes grow. Two factors explain this behavior: (i) a lack of statistical robustness to violation of the method's assumptions, as also observed in earlier tests; and (ii) interaction interpretation issues. These results confirm that ART is not scale-invariant, interpreting interactions relative to the scale of the observed data.

@@

3. The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations --- regardless of the method used. Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions. As previously noted, @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model. Our findings clearly show that this is not the case.

>>>>

- *RNK.* Error rates explode when both main effects exceed a certain level (e.g., when $a_1, a_2 \ge 4$ and $n = 20$). This problem arises from the way the rank transformation distorts interaction effects (see @fig-interactions-rank). For all continuous distributions, we observe that error rates exhibit the exact same patterns.

@@

- *ART.* It maintains correct error rates as long as population distributions are normal. For all other distributions, error rates increase rapidly as effect sizes grow. Two factors explain this behavior: (i) a lack of statistical robustness to violation of the method's assumptions, as also observed in our previous experiments; and (ii) interaction interpretation issues. These results confirm that ART is not scale-invariant, interpreting interactions relative to the scale of the observed data.

@@

3. The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations --- regardless of the method used. Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions. Note that @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model. Our findings clearly show that this is not the case.

### 2025-07-08 text/70-CaseStudies.qmd
small corrections

Like @Martin:2023, we investigate a $4 \times 2 \times 2$ repeated-measures design and set the number of participants to $n=22$. However, we now test a perfectly balanced design. We transform the latent variable to a Bernoulli distribution with a success probability parameter $p=.46$ ---average success rate found by @Martin:2023--- and then assess the Type I error rate of ART in the absence of any effect. We also assess the Type I error rate of regular ANOVA (LMER model) with no transformation (PAR). Our results over 5000 iterations are as follows:

>>>>

Like @Martin:2023, we investigate a $4 \times 2 \times 2$ repeated-measures design and set the number of participants to $n=22$. However, we now test a perfectly balanced design. We transform the latent variable to a Bernoulli distribution with a success probability parameter $p=.46$ --- average success rate found by @Martin:2023 --- and then assess the Type I error rate of ART in the absence of any effect. We also assess the Type I error rate of regular ANOVA (LMER model) with no transformation (PAR). Our results over 5000 iterations are as follows:

### 2025-07-08 text/80-Recommandations.qmd
small corrections

### Priviledging parametric methods
The scope of rank-based nonparametric statistics is very limited. @baguley2012 summarizes a fundamental concern regarding their use as follows:

@@

In our prior research [@fages:2022], we applied ordered probit models [@Burkner:2019;@Christensen2023ordinal] for analyzing ordinal responses in two experimental studies. Since the HCI community is not familiar with these models, we found it necessary to provide motivation for their use and offer additional explanations to reviewers and readers on how to interpret results. These models offer numerous advantages, including support for interval estimation and prediction in addition to null hypothesis testing. However, researchers are required to acquaint themselves with concepts of generalized mixed-effects models or Bayesian inference in order to use them correctly. They also come with certain limitations -- for example, the need to carefully specify priors to ensure convergence in certain scenarios, or their reliance on strict assumptions about populations variances and thresholds.

>>>>

### Privileging parametric methods
The scope of rank-based nonparametric statistics is very limited. @baguley2012 summarizes a fundamental concern regarding their use as follows:

@@

In our prior research [@fages:2022], we applied ordered probit models [@Burkner:2019;@Christensen2023ordinal] for analyzing ordinal responses in two experimental studies. Since the HCI community is not familiar with these models, we found it necessary to provide motivation for their use and offer additional explanations to reviewers and readers on how to interpret results. These models offer numerous advantages, including support for interval estimation and prediction in addition to null hypothesis testing. However, researchers are required to acquaint themselves with concepts of generalized mixed-effects models or Bayesian inference in order to use them correctly. These methods also come with certain limitations --- for example, the need to carefully specify priors to ensure convergence in certain scenarios, or their reliance on strict assumptions about population thresholds.

### 2025-07-08 text/abstract.qmd
small corrections

Surprisingly, ART functions as intended only under the strict assumption of a linear model with continuous responses, equal variances, and no missing data. In all other scenarios, it performs worse than simpler rank transformation methods and, in many cases -- such as with ordinal data -- significantly worse than parametric ANOVA. Our analysis of real-world data further demonstrates that using ART often leads to incorrect conclusions. By contrast, the inverse normal transformation proves to be the most versatile and robust nonparametric method, though we also identify scenarios where it inflates errors or distort effects.

>>>>

Surprisingly, ART functions as intended only under the strict assumption of a linear model with continuous responses, equal variances, and no missing data. In all other scenarios, it performs worse than simpler rank transformation methods and, in many cases --- such as with ordinal data --- significantly worse than parametric ANOVA. Our analysis of real-world data further demonstrates that using ART often leads to incorrect conclusions. By contrast, the inverse normal transformation proves to be the most versatile and robust nonparametric method, though we also identify scenarios where it inflates errors or distort effects.
